\chapter{Results}
\label{ch:results}
% - results and discussion
% - compare models and hyperparameters and feature sets and windows and attacks (easy vs stealthy performance)
% - And how it compares to IT


% plot correlation between parameters and certain feature sets and window lengthS?

% if an attacker could spend some time in the network they could observe the packets and try to match the metatada of his attack as much as possible

% possible big difference between test and val because of the misconfigurations in validation data.

\section{Evaluation Overview}
% Briefly explain models, splits, metrics and how to read the results.
\textcolor{blue}{is this too much repetition?}


This chapter presents the experimental results of the evaluated anomaly detection models on the constructed reconnaissance detection datasets. The goal of the evaluation is to determine to what extent unsupervised models trained only on benign conntrack-derived traffic can detect injected reconnaissance activity in real OT background traffic. As described in Section~\ref{sec:experimental_configuration}, the configurations vary in feature set, aggregation window length, anomaly threshold, and model-specific parameters.
For k-NN, the search space is more limited than for K-Means and the autoencoder
because of its higher computational cost. The k-NN results are therefore
interpreted as a targeted evaluation of validation-selected feature sets rather
than as a full sweep over all candidate feature sets.


The experiments are evaluated in two stages. The main model selection is performed on customer 1. First, the customer 1 validation split is used to compare configurations and identify promising model, feature-set and parameter combinations. These validation results are used for model selection and for understanding the effect of design choices such as window size and feature selection. The selected configurations are then evaluated on the customer 1 test split, which is kept separate from the selection process and is used to assess how well the chosen configurations generalize to another period of network traffic. In a final step, the selected top configurations are also evaluated on customer 2 as an additional cross-environment check.

The results are reported using both window-level and range-level metrics.
Window-level metrics treat each source-host window as an individual prediction
and therefore measure how accurately the model classifies separate time windows. Range-level metrics instead group consecutive anomalous windows into alarm ranges and compare these ranges with the true attack intervals. This provides a more operational view of the results, since reconnaissance activity usually unfolds over a period of time rather than in a single isolated window.

Both views are used throughout the discussion. Window-level F1 indicates how
precise the model is at the individual-window level, while range-level F1
indicates whether complete reconnaissance scenarios are detected as temporal
events and how many separate false alarm ranges are produced. A model may
therefore have a moderate window-level score while still detecting the relevant attack interval, or a strong window-level score while producing fragmented alarms.

Unless stated otherwise, the tables in this chapter report configurations selected according to validation performance, followed by their corresponding test-set performance.  The reported overall scores aggregate all evaluated reconnaissance scenarios, ranging from the more aggressive T5/T4 scans to the slower T1/T0 scans. Differences between validation and test results are interpreted as an indication of robustness and sensitivity to the underlying benign traffic period.



\section{Validation-Based Feature Set Selection}
\label{sec:validation_feature_selection}
% Show top validation feature sets, explain reduced k-NN search.

Before comparing the final model configurations on the test split, the validation results were used to identify the most promising feature sets. This step was especially important because the full feature-set search could not be repeated exhaustively for k-NN due to its higher computational cost, as discussed in Section~\ref{sec:knn_resource_constraints}. The selection was therefore based on the full K-Means and autoencoder validation sweeps, which covered the complete planned feature-set space. 

\definecolor{selectedrow}{RGB}{230,242,255}

\
\begin{table}[htbp]
\centering
\small
\renewcommand{\arraystretch}{1.25}
\setlength{\tabcolsep}{3.5pt}
\makebox[\textwidth][c]{%
\begin{tabularx}{1.05\textwidth}{>{\raggedright\arraybackslash}X l l l l}
\toprule
Feature set & \#configs & Mean RF1 & Mean F1 & Score \\
\midrule

\rowcolor{selectedrow}
\feat{golden\_minimal} 
& 896 & 0.224 & 0.261 & 0.242 \\

\rowcolor{selectedrow}
\feat{omni\_core\_clean} 
& 896 & 0.214 & 0.250 & 0.232 \\

\feat{adam\_entropy} 
& 336 & 0.183 & 0.225 & 0.204 \\

\rowcolor{selectedrow}
\feat{low\_slow\_clean} 
& 896 & 0.127 & 0.157 & 0.142 \\

\feat{T2\_stealth\_scan} 
& 896 & 0.099 & 0.116 & 0.107 \\

\feat{T3\_T4\_T5\_loud\_scan} 
& 896 & 0.116 & 0.085 & 0.101 \\

\feat{recon\_core\_clean\_temporal} 
& 896 & 0.075 & 0.090 & 0.083 \\

\feat{ablation\_entropy\_only} 
& 896 & 0.061 & 0.050 & 0.055 \\

\feat{recon\_core\_clean} 
& 896 & 0.050 & 0.059 & 0.054 \\

\feat{horizontal\_scan\_clean} 
& 336 & 0.072 & 0.035 & 0.054 \\

\feat{ablation\_asymmetry\_only} 
& 336 & 0.000 & 0.000 & 0.000 \\

\bottomrule

\end{tabularx}
}
\caption{Validation-based feature set ranking. RF1 denotes range-level F1, and \#configs is the number of validation configurations included for each feature set. The selection score is computed as the average of mean validation RF1 and mean validation window-level F1. Highlighted rows indicate the feature sets selected for the final cross-model comparison and targeted k-NN evaluation.}
\label{tab:validation_feature_set_ranking}
\end{table}



The goal of this selection was not only to find the highest individual validation score, but to identify feature sets that performed well across multiple reconnaissance scenarios. Feature sets that were designed for one specific attack type were therefore interpreted with caution, even when they achieved strong results on that specific attack. Instead, preference was given to feature sets with broader coverage across the evaluated attacks. To summarize the validation results across feature sets, a compact selection score is used. This score gives equal weight to the mean validation range F1 and the mean validation window-level F1. Table~\ref{tab:validation_feature_set_ranking} shows the resulting ranking. 


The number of configurations is not identical for all feature sets. The general-purpose feature sets were included in the full K-Means and autoencoder validation sweeps. In contrast, \feat{csv\_adam\_entropy} was evaluated as an ADAM-inspired k-NN baseline and therefore has fewer configurations. It will be further discussed in Section~\ref{sec:adam_baseline}. Similarly, \feat{csv\_horizontal\_scan\_clean} and \feat{csv\_ablation\_asymmetry\_only} were dropped earlier based on their weaker validation behavior, so their scores are based on a smaller set of configurations. 

The two strongest feature sets are \feat{csv\_golden\_minimal} and \feat{csv\_omni\_core\_clean}, which both achieve the highest combined validation scores. \feat{csv\_low\_slow\_clean} is also retained because it is the next strongest general-purpose feature set and is relevant for slower reconnaissance behaviour. Although \feat{csv\_T2\_stealth\_scan} achieves competitive scores in some cases, it is more closely tied to a specific attack type and is therefore not selected.

The three highest-scoring general feature sets thus are \feat{csv\_golden\_minimal}, \texttt{csv\_omni\_core\_clean}, and \texttt{csv\_low\_slow\_clean}. These are therefore retained for the final cross-model comparison and for the reduced k-NN evaluation.



% \subsection{ADAM}
% In addition to the feature sets constructed specifically for reconnaissance
% detection, an ADAM-inspired feature set was also evaluated. As introduced in Section~\ref{sec:literature_models}, this feature set was
% included because ADAM uses entropy-based traffic characteristics for anomaly
% detection in cyber-physical systems, and therefore provides a useful comparison
% point for evaluating whether a more compact entropy-oriented representation is
% sufficient for the present conntrack-based reconnaissance detection task.

% The ADAM-inspired feature set was evaluated using the same validation procedure
% as the other candidate feature sets. However, its validation performance was not
% competitive and hence dropped at an early stage, as shown in Table~\ref{tab:adam_feature_set_results}


% FIXXXXXXXXX TABLEEEEEEEEEEEEEEE

% These results suggest that the ADAM-inspired representation does not capture
% enough of the reconnaissance-specific behaviour present in the constructed
% dataset. In particular, the strongest selected feature sets combine several types
% of indicators, including destination spread, short-flow behaviour, failed or
% asymmetric communication, and temporal aggregation. The ADAM-inspired feature set
% is more limited in this respect, which may explain its weaker performance on the
% evaluated reconnaissance scenarios.

\section{Overall Model Performance}
% Best final configurations, validation + test table.
\label{sec:overall_model_performance}
\subsection{Validation Ranking Results}
The next step is to compare the best-performing configurations across the evaluated model families. Since many model, feature-set, window-size and threshold combinations were evaluated, the focus lies on the strongest configurations and on the main patterns that emerge from the validation results. The full top-ranked validation configurations are provided in Appendix~\ref{tab:appendix_top_val_f1} and Appendix~\ref{tab:appendix_top_val_rf1}, where the models are ranked separately by window-level F1 and range-level F1. The appendix rankings show that the highest window-level F1 scores are dominated by CPS-GUARD configurations with 60-second windows. These configurations achieve very high window-level F1 and also strong range-level scores. In contrast, several of the highest range-F1 configurations use 5-minute windows and obtain perfect or near-perfect range precision with high range recall.

\subsection{Interpreting Window-Level and Range-Level Metrics}
Both window-level F1 and range-level F1 are used because they capture different aspects of the detection task. Window-level F1 evaluates each source-host window as a separate prediction, and therefore measures how accurately the model labels individual time windows. Range-level F1 instead groups consecutive anomalous windows into alarm ranges and compares these with the true reconnaissance intervals. This gives a more operational view of the detector, since an analyst would usually investigate an alarm period rather than isolated individual windows.

The two metrics can lead to different conclusions. A model may achieve a high range-level score even when it only detects part of an attack interval, as long as at least one alarm range overlaps the true attack range. Conversely, a model may detect many attack windows but still obtain a lower range-level score if its false positives are spread over many separate alarm ranges. Larger aggregation windows and range-level evaluation also reduce the number of evaluation units, so a single missed attack range or false alarm range can have a strong effect on the final score.

The rankings also show why F1 should not be interpreted in isolation. Several configurations achieve similar F1 or range F1 values while making different trade-offs between precision and recall. In particular, a high range F1 can be obtained either by detecting almost all attack ranges with some additional false alarm ranges, or by producing very few false alarm ranges while missing one or more attack ranges. Although both types of error are relevant, missed attack ranges are more problematic in the context of early reconnaissance detection. A small number of additional false alarm ranges can still be inspected by an analyst, whereas a false negative means that an entire reconnaissance interval may remain unnoticed. Therefore, when configurations have comparable F1 scores, preference is given to models with no, or fewer, range false negatives. However, false positives remain important for operational usability, so the main ranking still uses F1 rather than a recall-weighted \(F_{\beta}\)-score.

\section{Selected Models}

\begin{table}[htbp]
\centering
\scriptsize
\renewcommand{\arraystretch}{1.15}
\setlength{\tabcolsep}{2pt}
\begin{adjustbox}{width=1.35\textwidth,center}
\begin{tabular}{l l p{0.22\textwidth} l l r r r r r r r r}
\toprule
Model & Basis & Feature set & Win. & Parameters
& Val R & \textbf{Val F1}
& Val RP & Val RR & \textbf{Val RF1}
& Test F1 & Test RF1 \\
\midrule

\rowcolor{selectedrow}
CPS-GUARD
& F1
& \feat{golden\_minimal}
& 60 s
& 36-6-36, cont. $10^{-3}$
& 1.000 & \textbf{0.996}
& 0.830 & 1.000 & \textbf{0.907}
& 0.995 & 1.000 \\

CPS-GUARD
& RF1
& \feat{omni\_core\_clean}
& 5 min
& 64-16-64, cont. $10^{-5}$
& 0.385 & \textbf{0.556}
& 1.000 & 0.933 & \textbf{0.966}
& 0.066 & 0.500 \\

\rowcolor{selectedrow}
K-Means
& F1
& \feat{golden\_minimal}
& 5 min
& $c=10$, $q=0.999$
& 0.981 & \textbf{0.937}
& 0.547 & 0.967 & \textbf{0.699}
& 0.875 & 0.206 \\

K-Means
& RF1
& \feat{golden\_minimal}
& 5 min
& $c=5$, $q=0.999999$
& 0.385 & \textbf{0.548}
& 0.875 & 0.933 & \textbf{0.903}
& 0.066 & 0.500 \\

\rowcolor{selectedrow}
k-NN
& F1
& \feat{golden\_minimal}
& 60 s
& $k=10$, $q=0.9999$
& 1.000 & \textbf{0.975}
& 0.520 & 1.000 & \textbf{0.684}
& 0.856 & 0.080 \\

k-NN
& RF1
& \feat{low\_slow\_clean}
& 5 min
& $k=10$, $q=0.9999$
& 0.390 & \textbf{0.561}
& 1.000 & 0.933 & \textbf{0.966}
& 0.097 & 0.615 \\

\rowcolor{selectedrow}
ADAM-inspired k-NN
& F1/RF1
& \feat{adam\_entropy}
& 1 h
& $k=1$, $q=0.99999$
& 0.829 & \textbf{0.906}
& 1.000 & 0.714 & \textbf{0.833}
& 0.906 & 0.833 \\
\bottomrule
\end{tabular}
\end{adjustbox}
\caption{Selected validation candidates per model family. The highlighted rows indicate the representative configurations retained for the main comparison. F1 denotes window-level F1, RP/RR/RF1 denote range-level precision, recall and F1.}
\label{tab:selected_model_candidates}
\end{table}

Based on this interpretation, Table~\ref{tab:selected_model_candidates} summarizes the strongest customer 1 validation candidates per model family under the two ranking criteria. For each model familiy, the table includes the best configuration when ranked by range-level F1 and by window-level F1. All configurations are evaluated on the same validation attack and test scenarios. The reported metrics are therefore computed against the same ground truth attack windows and attack ranges. 

For CPS-GUARD and k-NN, the F1-ranked configuration is retained because it achieves perfect validation range recall, meaning that no validation attack ranges are missed. The RF1-ranked alternative obtains higher range-level F1, but this is mainly due to fewer false alarm ranges and comes at the cost of missed attack ranges. 

For K-Means, the same trade-off is visible. The RF1-ranked configuration has higher range precision and range F1, but misses more attack ranges and has much lower window-level recall. The F1-ranked configuration is therefore retained as the representative K-Means candidate, although its lower range precision indicates that it produces more false alarm ranges. 

For k-NN, the F1-ranked configuration is also retained as the main representative configuration. Although the RF1-ranked k-NN configuration obtains a higher validation range-level F1, it does so with lower window-level recall and with missed attack ranges. The RF1-ranked k-NN configuration reaches a higher validation range-level F1, but it misses attack ranges and has substantially lower window-level recall. The F1-ranked configuration detects all validation attack ranges and achieves perfect validation recall at the window level, although this comes with lower range precision.

The ADAM-inspired entropy configuration performs competitively, however, it does not achieve perfect range recall and therefore misses some attack ranges. 




\subsection{Test Performance of Selected Models}

After selecting the representative configurations from the customer 1 validation results,
their performance is evaluated on the test split. Table~\ref{tab:selected_test_performance}
reports the test-set precision, recall and F1-score at both the window level and
the range level.

\begin{table}[htbp]
\centering
\scriptsize
\renewcommand{\arraystretch}{1.15}
\setlength{\tabcolsep}{3pt}
\begin{adjustbox}{width=1.35\textwidth,center}
\begin{tabular}{l p{0.23\textwidth} l l r r r r r r }
\toprule
Model & Feature set & Win. & Parameters
& Test P & Test R & \textbf{Test F1}
& Test RP & Test RR & \textbf{Test RF1} \\
\midrule

CPS-GUARD
& \feat{golden\_minimal}
& 60 s
& 36-6-36, cont. $10^{-3}$
& 1.000 & 0.989 & \textbf{0.995}
& 1.000 & 1.000 & \textbf{1.000} \\

K-Means
& \feat{golden\_minimal}
& 5 min
& $c=10$, $q=0.999$
& 0.791 & 0.980 & \textbf{0.875}
& 0.119 & 0.778 & \textbf{0.206} \\

k-NN
& \feat{golden\_minimal}
& 60 s
& $k=10$, $q=0.9999$
& 0.750 & 0.996 & \textbf{0.856}
& 0.04 & 0.818 & \textbf{0.080} \\

ADAM-inspired k-NN
& \feat{adam\_entropy}
& 1 h
& $k=1$, $q=0.99999$
& 1.000 & 0.829 & \textbf{0.906}
& 1.000 & 0.714 & \textbf{0.833} \\


\bottomrule
\end{tabular}
\end{adjustbox}
\caption{Test-set performance of the validation-selected representative configurations. P/R/F1 denote window-level precision, recall and F1. RP/RR/RF1 denote range-level precision, recall and F1. }
\label{tab:selected_test_performance}
\end{table}

% confusion matrix of best feature set per attack type?


The selected CPS-GUARD configuration generalizes best to the test split. It achieves a window-level F1-score of 0.995 and detects all test attack ranges, resulting in a range-level F1-score of 1.000. This indicates that the validation performance was not only caused by overfitting to the validation period.It is also notable that this configuration uses the \feat{golden\_minimal} feature set and the original CPS-GUARD-style 36-6-36 architecture.
% best architecture is the one suggested by original paper. Also golden_minimal best preforming feature set.

The selected K-Means configuration retains strong window-level recall on the test split, but its range-level precision is low. This means that although many attack windows are detected, the predictions are fragmented into several false alarm ranges, which would increase the amount of analyst work in an operational setting.

The ADAM-inspired entropy baseline also transfers consistently to the test split, with high precision and a window-level F1-score of 0.906. However, its range recall remains lower than CPS-GUARD because two attack ranges are missed.


\subsection{Computational Cost and Practical Trade-offs}
\label{sec:computational_tradeoffs}

Besides detection performance, the models also differ in computational cost. All evaluated model families are feasible for real-time use in the sense that their evaluation time remains below the smallest evaluated window length of 60 seconds. However, their efficiency differs substantially. 

K-Means is the fastest model family, because each new window only has to be compared with a small number of centroids. CPS-GUARD is more expensive because each window must pass through the autoencoder, but inference still has a fixed cost once the model is trained. k-NN is the slowest option, since each evaluated window must be compared with the stored benign reference windows to compute nearest-neighbour distances. This cost becomes especially relevant for short windows, where both the number of reference windows and the number of evaluated windows are larger.






\subsection{Detection Performance Across Attack Types}
Table~\ref{tab:per_attack_selected_models} shows that the selected CPS-GUARD configuration is the only model that performs consistently well across all evaluated reconnaissance scenarios. It detects all test attack ranges and achieves near-perfect or perfect window-level F1 for every attack type. This indicates that its overall performance is not driven by only one easy attack scenario.

\begin{table}[htbp]
\centering
\scriptsize
\renewcommand{\arraystretch}{1}
\setlength{\tabcolsep}{8pt}
\begin{adjustbox}{width=1\textwidth,center}
\begin{tabular}{l l r r r r r r r r}
\toprule
Model & Metric & T1/T0 & T2 & T5/T4/T3 & APT \\
\midrule
CPS-GUARD & F1  & 1.000 & 0.994 & 1.000 & 1.000 \\
CPS-GUARD & RF1 & 1.000 & 1.000 & 1.000 & 1.000 \\

K-Means & F1  & 0.000 & 0.962 & 0.513 & 0.696 \\
K-Means & RF1 & 0.000 & 0.133 & 0.316 & 0.316 \\

k-NN & F1  & 0.016 & 0.955 & 0.379 & 0.655 \\
k-NN & RF1 & 0.036 & 0.038 & 0.105 & 0.136 \\

ADAM-inspired k-NN & F1  & 0.000 & 0.955 & 1.000 & 0.857 \\
ADAM-inspired k-NN & RF1 & 0.000 & 1.000 & 1.000 & 1.000 \\
\bottomrule
\end{tabular}
\end{adjustbox}
\caption{Test performance per reconnaissance scenario for the selected configurations. F1 denotes window-level F1 and RF1 denotes range-level F1.}
\label{tab:per_attack_selected_models}
\end{table}

The other models show more scenario-dependent behaviour. K-Means and k-NN both detect parts of the T2, T5/T4/T3 and APT scenarios, but their range-level scores are much lower because their false positives are spread over many separate alarm ranges. The ADAM-inspired entropy baseline is very precise on the scenarios it detects, but it misses the T1/T0 scenario completely, showing that the entropy-only representation is less robust across different reconnaissance styles.

% Temporal granularity trade-offs.

% \section{Influence of Feature Set}
% % What feature families worked and why.

% \section{Detection Performance per Reconnaissance Scenario}
% % Loud scans, T2, T1/T0, low-and-slow.

\section{Cross-Environment Evaluation on Customer 2}
\label{sec:customer2_cross_environment}
% while the strongest difference in features is the same for customer 2, the variance in values is much bigger for customer 2 which explains more FP.


\section{Operational Stuff}
% SOC usefulness, false alarms, recommended configuration.
% Optional short IT comparison here if you have it.

From an operational perspective, the selected CPS-GUARD configuration is the most suitable candidate among the evaluated models. It preforms well across both evaluation metrics, meaning that. it not only identifies many malicious windows but also produces alarm ranges that correspond well to the injected reconnaissance intervals. 

This shows why range-level evaluation is important for practical use. A detector that achieves high window-level recall can still be difficult to use if its false positives are spread over many separate alarm ranges. For K-Means and k-NN, while they detect many attack windows, they produce lower range precision. In a SOC setting, this would increase the number of alarm periods that an analyst has to inspect. Therefore, the most useful configuration is not necessarily the one with the highest window-level recall, but the one that provides a manageable number of meaningful alarm ranges.

sht about feature sets and how they are environment specific and customer 2 :d.






% \section{Overall Detection Performance}
% Compare best k-NN, K-Means and CPS-GUARD-style autoencoder configurations.

% \section{Influence of Window Size}
% Discuss 1 min, 5 min, 1 h, 5 h trade-offs.

% \section{Influence of Feature Set}
% Discuss which feature families were useful and whether compact sets were enough.

% \section{Detection Performance per Reconnaissance Scenario}
% Discuss T5/T4/T3, T2, T1/T0, low-and-slow.

% \section{Comparison with IT Traffic}
% Discuss whether results transfer outside OT.

% \section{Operational Discussion}
% Discuss deployment meaning, false alarms, SOC usefulness, recommended setup.

% \section{Limitations}
% Discuss injected attacks, limited environments, noisy benign data, validation dependence.